---
title: "lab3"
format: html
---

class notes:

stacked area plot

- good at minimizing space
- bad at getting at exact values and how different things are relative to each other

stacked bar plot:

- good at minimizing space use
- bad at getting at exact values and how different things are relative to each other

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#..........................load packages.........................
library(tidyverse)

#..........................import data...........................
tuesdata <- tidytuesdayR::tt_load('2021-07-20')
drought <- tuesdata$drought

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                       data vis: in lab exercise                          ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# subset
drought_subset <- drought %>%
  filter(state_abb == c("CA", "OR", "WA"))

# make a plot

drought_subset |> 
  ggplot(drought_subset, mapping = aes(x = area_pct, y = state_abb, fill = after_stat(x))) +
  ggridges::geom_density_ridges_gradient() +
  scale_fill_gradientn(colors = c("#2C5374","#849BB4", "#D9E7EC", "#EF8080", "#8B3A3A"))


##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                            wrangle drought data                          ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

drought_clean <- drought |>

  # select cols of interest & update names as needed ----
# this wrangling is important because: easier to catch mistakes, less data to deal with
  select(date = valid_start, state_abb, drought_lvl, area_pct) |> 

  # add year, month & day cols using {lubridate} fxns ----
# this wrangling is important because: we want to plot by/ group by year
  mutate(year = year(date),
         month = month(date, label = TRUE, abbr = TRUE),
         day = day(date)) |>

  # add drought level conditions names ----
# this wrangling is important because: aesthetics/ logical order
  mutate(drought_lvl_long = factor(drought_lvl,
                            levels = c("D4", "D3", "D2", "D1","D0", "None"),
                            labels = c("(D4) Exceptional", "(D3) Extreme",
                                       "(D2) Severe", "(D1) Moderate", "(D0) Abnormally Dry", 
                                       "No Drought"))) |>
  
  # reorder cols ----
  select(date, year, month, day, state_abb, drought_lvl, drought_lvl_long, area_pct)

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##       create stacked area plot of CA drought conditions through time     ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

# this wrangling is important because: 

drought_clean |> 
  
  # remove drought_lvl "None" & filter for just CA ----
# this wrangling is important because: smaller dataset easier to assess trends, and faster rendering speed
  filter(drought_lvl != "None",
         state_abb == "CA") |> 
  
  # create ggplot ----
  ggplot(mapping = aes(x = date, y = area_pct, fill = drought_lvl_long)) +
  
  # reverse order of groups so level D4 is closest to x-axis ----
  geom_area(position = position_stack(reverse = TRUE)) +
  
  # update colors to match US Drought Monitor (colors selected using ColorPick Eyedropper from original USDM data viz) ----
  scale_fill_manual(values = c("#853904", "#FF0000", "#FFC100", "#FFD965", "#FFFF00")) +
  
  # set x-axis breaks & remove padding between data and x-axis ----
  scale_x_date(breaks = scales::breaks_pretty(n = 10),
               expand = c(0, 0)) +

  # set y-axis breaks & convert values to percentages & & remove padding between data and y-axis----
  scale_y_continuous(breaks = seq(0, 100, by = 10),
                     labels = scales::label_percent(scale = 1),
                     expand = c(0, 0)) +
  
  # add title ----
  labs(title = "Drought area in California") 

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##            BONUS: using {geofacet} to plot data for all states           ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#......remove HI & AK (no data) and also DC from preset grid.....
mygrid <- geofacet::us_state_grid1 |>
  filter(!code %in% c("DC", "HI", "AK"))

#..............................plot..............................

# NOTE: this takes a minute to render!

drought_clean |> 
  filter(drought_lvl != "None") |> 
  ggplot(aes(x = date, y = area_pct, fill = drought_lvl_long)) +
  geom_area(position = position_stack(reverse = TRUE)) +
  geofacet::facet_geo(~state_abb, grid = mygrid) +
  scale_fill_manual(values = c("#853904", "#FF0000", "#FFC100", "#FFD965", "#FFFF00"))
```

```{r}
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                                    setup                                 ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#.........................load libraries.........................
library(tidycensus)
library(tidyverse)
library(janitor)
library(gghighlight)

#.........................source API key.........................
source(here::here("week3", "KEYS.R"))
census_api_key(censusKEY)

#..........................import data...........................
lyme <- read_csv(here::here("week3", "data", "LD-Case-Counts-by-County-01-20.csv"))

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                          wrangle lyme disease data                       ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#............wide to long (plus some other wrangling)............
lyme_clean <- lyme |> 
  
  # make col names snake_case ----
  janitor::clean_names() |> 
  
  # rename columns ----
  rename(city = ctyname, state = stname, status = ststatus) |> 
  
  # wide to long (tidy) years
  pivot_longer(cols = 6:25, names_to = "city_year", values_to = "reported_cases") |> 
  
  # remove "cases" from the year & coerce year from chr to factor ----
  mutate(year = str_remove(city_year, pattern = "cases"),
         year = as.factor(year)) |> 
  
  # select necessary cols ----
  select(year, city, state, status, reported_cases)

#................calculate total cases per state.................
lyme_by_state <- lyme_clean |> 
  group_by(year, state) |> 
  summarize(total_cases = sum(reported_cases)) 

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                      request / wrangle population data                   ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

#...................get pop estimates by state...................
us_state_pop <- get_estimates(geography = "state", 
                              product = "population",
                              state = NULL, 
                              year = 2019) |> 
  filter(variable == "POP") |> 
  select(state = NAME, population = value) 

#........................write data to csv.......................
# optional, but recommended in case you want to work offline, the API is down, etc. (you can then read in your saved data file rather than run the above code)
# write_csv(us_state_pop, file = here::here("week3", "data", "us_state_pop.csv"))

##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
##                            join lyme & pop dfs                           ----
##~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~

lyme_pop <- left_join(lyme_by_state, us_state_pop) |> 
  
  # add col with num of 100k people per state ----
  mutate(pop100k = population/100000) |> 
  
  # calculate num cases per 100k people (common way of reporting disease incidence) ----
  mutate(cases_per100k = total_cases/pop100k) 
```

```{r}
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state)) + 
  geom_line()

lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state)) + 
  geom_line() +
  geom_point()

# just new jersey
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state)) +
  geom_line() +
  gghighlight::gghighlight(state == "New Jersey") 

# highlight only the ones with more than 100, "at some point thei spiked above 100k"
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state, color = state)) + # group is always needed for a line plot
  geom_line() +
  gghighlight::gghighlight(max(cases_per100k) > 100) # note all the conditional sin gghighlight
```


# play with aspect ratio
```{r}
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  filter(state == "Vermont") |> 
  ggplot(aes(x = year, y = cases_per100k, group = state)) +
  geom_line() +
  scale_y_continuous(breaks = seq(0, 190, by = 1)) # set breaks at each 1 unit on the y axis # this allows us to understand the difference in the magnitude of our units on both axes

lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  filter(state == "Vermont") |> 
  ggplot(aes(x = year, y = cases_per100k, group = state)) +
  geom_line() +
  scale_y_continuous(breaks = seq(0, 190, by = 1)) +
  coord_fixed(ratio = 1) # now aspect ratio of 1


lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  filter(state == "Vermont") |> 
  ggplot(aes(x = year, y = cases_per100k, group = state)) +
  geom_line() +
  scale_y_continuous(breaks = seq(0, 190, by = 1)) +
  coord_fixed(ratio = 1/50) # now aspect ratio of 1/50

# last cut out the tick marks
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  filter(state == "Vermont") |> 
  ggplot(aes(x = year, y = cases_per100k, group = state)) +
  geom_line() +
  #scale_y_continuous(breaks = seq(0, 190, by = 1)) +
  coord_fixed(ratio = 1/50) 

```
# area chart 1 group

the only thing that matters is the top line, the fill in doesnt specify anything with one variable!

```{r}
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  filter(state == "New Jersey") |> # "any observations that are equal equal to"
  ggplot(aes(x = year, y = cases_per100k, group = state)) +
  geom_area() 

# Expand area to panel margins
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  filter(state == "New Jersey") |> 
  ggplot(aes(x = year, y = cases_per100k, group = state, fill = state)) +
  geom_area() +
  scale_fill_manual(values = c("#047C90")) +
  scale_x_discrete(expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0)) +
  theme(legend.position = "none")

```
# area chart mult groups

these are stacked on top of each other, need todo subtraction to get at each state's. Top line is totalincidence for all beneath it. 

helpful when you want to show the evolution of a whole, but still weird...proportional stacked area chart is recommended

```{r}
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  filter(state %in% c("Maine", "Rhode Island", "New Hampshire", "Vermont")) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state, fill = state)) +
  geom_area()

# Proportional stacked area charts plot percentage contribution instead of absolute numbers on the y-axis
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  filter(state %in% c("Maine", "Rhode Island", "New Hampshire", "Vermont")) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state, fill = state)) +
  geom_area(position = "fill") +
  scale_y_continuous(labels = scales::label_percent(scale = 100))
```
# can use lines instead
```{r}
lyme_pop |> 
  filter(year %in% c(2010:2020)) |> 
  filter(state %in% c("Maine", "Rhode Island", "New Hampshire", "Vermont")) |> 
  ggplot(aes(x = year, y = cases_per100k, group = state, color = state)) +
  geom_line() +
  facet_wrap(~state)

```


